---
title: "Raster data analysis in R"
author: "Krzysztof Dyba"
format:
  html:
    toc: true
    toc-title: " "
    embed-resources: true
---

```{r}
#| message: false
#| warning: false

# load the package
library("terra")
```

# Introduction

In the `landsat` folder, there are 4 raster files with the `.tif` extension
that show the Warsaw area. This data comes from the [Landsat 9](https://science.nasa.gov/mission/landsat/)
satellite and represents selected ranges of the electromagnetic spectrum
recorded by the sensor.

| Band Number | Band Name                      | Wavelength (µm) |
|-------------|--------------------------------|-----------------|
| Band 2      | Blue                           | 0.45 – 0.51     |
| Band 3      | Green                          | 0.53 – 0.59     |
| Band 4      | Red                            | 0.64 – 0.67     |
| Band 5      | Near Infrared (NIR)            | 0.85 – 0.88     |

: Selected spectral bands from Landsat 9. {#tbl-landsat}

The purpose of this analysis is to assess the state of vegetation in this area
using simple thresholding and classification rules based on a spectral index.

We will start by listing the files we want to load using the `list.files()`
function. We need to specify the path to the folder and the file extension
using regular expression (`pattern = "\\.TIF$"`). In addition, we use the
`full.names` argument to return the full file paths.

```{r}
# list files from a directory
files = list.files("landsat", pattern = "\\.TIF$", full.names = TRUE)
files
```

Now that we have the files listed, we can load them using the `rast()` function.

```{r}
# load rasters
landsat = rast(files)
```

In the next step, we will inspect the basic metadata.

```{r}
# print metadata
landsat
```

For convenience, we can modify the default names of bands (layers) by referring
to the `names` attribute.

```{r}
# rename the layers
names(landsat) = c("Blue", "Green", "Red", "NIR")
# names(landsat) = c("B02", "B03", "B04", "B05")
```

Obviously, we can visualize this data using the simple `plot()` function.
By default, the viridis color palette is used. However, by using the `col`
argument, we can use a different one, such as shades of gray from `gray.colors()`.
For instance, let's display the NIR band -- we can select its name or index
using square brackets.

```{r}
#| layout-ncol: 2

plot(landsat[["NIR"]], main = "NIR Band")

colors = gray.colors(n = 20) # define the custom color scheme
plot(landsat[["NIR"]], main = "NIR Band", col = colors)
```

# Raster Operations

## Statistical Summaries

To calculate basic statistics, the easiest method is to use the `summary()`
function, which by default performs calculations on a sample of 100,000 raster
cells.

```{r}
#| warning: false

summary(landsat)
```

The statistical values above are not entirely correct. The proper range for
reflectance is from 0 to 1, where 0 indicates no reflection and 1 indicates
full reflection. The reason for this is that satellite data is typically stored
as integers rather than floating-point numbers, in order to reduce file size.

## Calculations

Therefore, we need to perform a simple scaling using scale factor ($0.0000275$)
and offset ($-0.2$), according to the following formula:

$$
x = x * \text{scale factor} + \text{offset}
$$

Mathematical operations on rasters are performed in exactly the same way as on
matrices. The formula will be automatically applied to all cells in the raster
and all layers of the raster.

```{r}
# scaling values
landsat = landsat * 0.0000275 - 0.2
```

Let's check the statistics based on all raster cells.

```{r}
#| warning: false

summary(landsat, size = Inf)
```

We can see that some values still fall outside the expected range, i.e. below
0 or above 1. These are outliers. There are several ways to remove them, but
the most common is the `clamp()` function. It requires setting lower and
upper threshold values, and depending on the `values` argument, outliers can
either be overwritten with the threshold value (`values = TRUE`) or removed
(`values = FALSE`). Typically, such values are removed, but if a large number
of observations are lost, they are overwritten instead.

```{r}
# method 1 (overwrite)
landsat = clamp(landsat, lower = 0, upper = 1, values = TRUE)
```

```{r}
#| eval: false

# method 2 (remove)
# landsat = clamp(landsat, lower = 0, upper = 1, values = FALSE)
```

## Multiband Visualization

In addition to single-layer visualizations, we can create more advanced
compositions consisting of three bands. For instance, a composition in
natural colors that consists of red, green, and blue bands. The `plotRGB()`
function is used for this, and we need to define the appropriate bands as
its arguments. In practice, it is usually necessary to stretch the contrast,
which can be done by setting the `stretch = "lin"` argument.

```{r}
# RGB composition
plotRGB(landsat, 3, 2, 1, stretch = "lin")
```

Using other spectral bands, we can create compositions in false colors.

## Raster Algebra

Let’s now proceed to assess vegetation condition. To do this, we will use a
spectral index called the normalized difference vegetation index (NDVI).
It is a simple indicator based on the red and near-infrared bands and it
is used to measure plant health and biomass levels. Its formula is as follows:

$$
\text{NDVI} = \frac{\text{NIR} - \text{Red}}{\text{NIR} + \text{Red}}
$$

Let's write this formula as a function.

```{r}
# define a function to calculate NDVI
calculate_ndvi = function(nir, red) {
  ndvi = (nir - red) / (nir + red)
  return(ndvi)
}
```

Next, let's call this function by specifying the appropriate bands.

```{r}
ndvi = calculate_ndvi(landsat[["NIR"]], landsat[["Red"]])
```

We obtained a result that falls within the range of -1 to 1. The higher the
value, the healthier the plants and the higher the biomass level. Low values
around 0 indicate bare soil, artificial surfaces, or water bodies.

Let's now create visualizations using a custom color palette ranging from
red (low values) to green (high values). We can generate such a palette with
the `colorRampPalette()` function and specifying the main colors and the
number of interpolated intervals between them.

```{r}
# divergent color palette
ndvi_colors = colorRampPalette(c("#8B0000", "#F5DEB3", "#006400"))
plot(ndvi, main = "NDVI", col = ndvi_colors(20), range = c(-1, 1))
```

Forests and croplands are marked in green, while urban areas and bodies of
water are marked in yellow.

## Thresholding

We can use a threshold on the NDVI raster to create a simple binary mask
showing areas covered by vegetation. We will use 0.4 as the threshold value.

```{r}
veg = ndvi > 0.4
plot(veg, col = c("white", "darkgreen"), main = "Vegetation")
```

## Simple Classification

A more complex approach is based on classification rules. To reclassify raster
values based on conditions, we use the `classify()` function. It requires
a matrix consisting of three columns. The first and second columns contain
the starting and ending range of raster values, and the third column contains
the new category value (ID).

```{r}
# reclassification matrix
rcl = matrix(
  c(-1.0, 0.2, 1,   # Non-vegetated / Water
     0.2, 0.4, 2,   # Sparse vegetation
     0.4, 0.6, 3,   # Moderate vegetation
     0.6, 1.0, 4),  # Dense vegetation
  ncol = 3, byrow = TRUE
)
```

```{r}
# raster value reclassification
ndvi_class = classify(ndvi, rcl)
```

The result is a new raster consisting of IDs from 1 to 4. Let's replace these
IDs with category names. We need to create a data frame with two columns.
The first column contains IDs, and the second contains labels. Then, using
the `levels()` function, we can assign these labels to their corresponding IDs.

```{r}
df = data.frame(
  id = unique(ndvi_class),
  labels = c("Non-vegetated", "Sparse", "Moderate", "Dense")
)

# assign labels to IDs 
levels(ndvi_class) = df
```

Let's take a look at how these categories look on the map.

```{r}
col_class = c("#e2e8f0", "#fefae0", "#78c679", "#065f46")
plot(ndvi_class, col = col_class)
```

We can also count the number of occurrences of each category using the `freq()`
function.

```{r}
# frequency table
freq(ndvi_class, bylayer = FALSE)
```

## Focal Operations

The created map shows a "salt-and-pepper" effect, in which incorrect categories
appear on homogeneous areas. This is due to the limitations of remote sensing
measurements. This can be corrected by using focal operations, which analyze
the nearest neighborhood of cells. In our case, we can use a modal filter in
`focal()` function that selects the most frequently appearing categories within
a moving window of a specified size. Let's assume the window size is 5 x 5 cells.

```{r}
# map smoothing using modal filter
ndvi_class_smooth = focal(ndvi_class, w = 5, fun = "modal", expand = TRUE)
```

Let's look at the smoothed map.

```{r}
plot(ndvi_class_smooth, col = col_class)
```

## Exporting Results

We can save the results of our analysis as a raster using the `writeRaster()`
function and specifying the save path along with the file extension.
GeoTIFF files are saved with LZW compression by default.

```{r}
#| eval: false

# save raster to disk
writeRaster(ndvi, "NDVI.tif")
```

For a raster with categories, we can define the data type as byte
(`datatype = "INT1U"`) instead of a floating point number, which will
significantly reduce the disk space occupied.

```{r}
#| eval: false

writeRaster(ndvi_class, "NDVI_class.tif", datatype = "INT1U")
```

# Task

1. Create a false color composition consisting of the following bands in this
order: "NIR", "Red", "Green".
2. Calculate the normalized difference water index (NDWI) for water detection.
Next, use thresholding to delineate the water areas and save the result to
disk in TIF format. To calculate NDWI, use the following formula:

$$
\text{NDWI} = \frac{\text{Green} - \text{NIR}}{\text{Green} + \text{NIR}}
$$

