---
title: "Raster data analysis in R"
author: "Krzysztof Dyba"
format:
  html:
    toc: true
    toc-title: " "
    embed-resources: true
---

```{r}
#| message: false
#| warning: false

# load the package
library("terra")
```

# Introduction

In the `data/landsat` folder, there are 4 raster files with the `.TIF` extension
that show the Warsaw area. These data come from the [Landsat 9](https://science.nasa.gov/mission/landsat/)
satellite and represent selected ranges of the electromagnetic spectrum
recorded by the sensor.

| Band Number | Band Name                      | Wavelength (µm) |
|-------------|--------------------------------|-----------------|
| Band 2      | Blue                           | 0.45 – 0.51     |
| Band 3      | Green                          | 0.53 – 0.59     |
| Band 4      | Red                            | 0.64 – 0.67     |
| Band 5      | Near Infrared (NIR)            | 0.85 – 0.88     |

: Selected spectral bands from Landsat 9. {#tbl-landsat}

The purpose of this analysis is to assess the state of vegetation in this area
using simple thresholding and classification rules based on a spectral index.

We will start by listing the files we want to load using the `list.files()`
function. We need to specify the path to the folder and the file extension
using a regular expression (`pattern = "\\.TIF$"`). In addition, we use the
`full.names` argument to return the full file paths.

```{r}
# list files from a directory
files = list.files("data/landsat", pattern = "\\.TIF$", full.names = TRUE)
files
```

Now that we have the files listed, we can load them using the `rast()` function.

```{r}
# load rasters
landsat = rast(files)
```

In the next step, we will inspect the basic metadata.

```{r}
# print metadata
landsat
```

For convenience, we can modify the default names of bands (layers) by referring
to the `names` attribute.

```{r}
# rename the layers
names(landsat) = c("Blue", "Green", "Red", "NIR")
# names(landsat) = c("B02", "B03", "B04", "B05")
```

Obviously, we can visualize this data using the simple `plot()` function.
By default, the viridis color palette is used. However, we can specify a
different palette using the `col` argument, such as shades of gray from
`gray.colors()`. For instance, let's display the NIR band -- we can select
it by name or index using double square brackets.

```{r}
#| layout-ncol: 2

plot(landsat[["NIR"]], main = "NIR Band")

colors = gray.colors(n = 20) # define the custom color scheme
plot(landsat[["NIR"]], main = "NIR Band", col = colors)
```

# Raster Operations

## Statistical Summary

To calculate basic statistics, the easiest method is to use the `summary()`
function, which by default performs calculations on a sample of 100,000 raster
cells.

```{r}
#| warning: false

summary(landsat)
```

The statistical values above do not reflect actual physical reflectance values.
The proper range for reflectance is from 0 to 1, where 0 indicates no reflection
and 1 indicates full reflection. The reason for this is that satellite data is
typically stored as 16-bit unsigned integers (values ranging from 0 to 65,535)
rather than floating-point numbers, in order to reduce file size.

## Data Preprocessing

Therefore, we must apply a scaling factor and an offset according to the
appropriate formula. Depending on the product or sensor, different formulas
and parameter values are used. For our dataset, the scaling factor is $0.0000275$
and the offset is $-0.2$. The formula is as follows:

$$
x' = x * \text{scale factor} + \text{offset}
$$

Mathematical operations on rasters are performed in exactly the same way as on
matrices. The formula will be automatically applied to all cells in the raster
and all layers of the raster.

```{r}
# scaling values
scale_factor = 0.0000275
offset = -0.2
landsat = landsat * scale_factor + offset
```

Let's check the statistics based on all raster cells (note that for large
rasters, processing all cells may be slow and memory intensive).

```{r}
#| warning: false

summary(landsat, size = Inf)
```

We can see that some values still fall outside the expected reflectance range
of 0 to 1. They may result from atmospheric correction artifacts, sensor noise,
measurement errors or outliers. There are several ways to handle them, but
the most common is the `clamp()` function. It requires setting lower and
upper threshold values, and depending on the `values` argument, outliers can
either be overwritten with the threshold value (`values = TRUE`) or removed
(`values = FALSE`). Typically, such values are removed, but if a large number
of observations are lost, they are overwritten instead.

```{r}
# method 1 (overwrite)
landsat = clamp(landsat, lower = 0, upper = 1, values = TRUE)
```

```{r}
#| eval: false

# method 2 (remove)
# landsat = clamp(landsat, lower = 0, upper = 1, values = FALSE)
```

## Multiband Visualization

In addition to single-layer visualizations, we can create more advanced
compositions consisting of three bands. For instance, a natural color
composition consists of the red, green, and blue bands. The `plotRGB()`
function is used for this, and we need to define the appropriate bands as
its arguments. Since red is the third layer, green is the second, and blue
is the first, we pass the values 3, 2, 1 as arguments to match these bands.
In practice, it is usually necessary to stretch the contrast, which can be
done by setting the `stretch = "lin"` argument.

```{r}
# RGB composition
plotRGB(landsat, r = 3, g = 2, b = 1, stretch = "lin")
```

Using other spectral bands, we can create compositions in false colors.
For example, a composition of NIR, Red, and Green bands highlights healthy
vegetation in bright red (due to its high NIR reflectance). Meanwhile, water
bodies appear black because they absorb NIR light, and built-up areas appear
as light blue or gray. This makes it much easier to distinguish different
land cover types than in natural colors.

## Vegetation Index

Let’s now proceed to assess the condition of the vegetation. To do this,
we will use a spectral index called the normalized difference vegetation
index (NDVI). It is a simple indicator based on the red and near-infrared
bands and it is used to indirectly measure plant health and biomass levels.
Its formula is as follows:

$$
\text{NDVI} = \frac{\text{NIR} - \text{Red}}{\text{NIR} + \text{Red}}
$$

Let's write this formula as a function.

```{r}
# define a function to calculate NDVI
calculate_ndvi = function(nir, red) {
  ndvi = (nir - red) / (nir + red)
  names(ndvi) = "NDVI"
  return(ndvi)
}
```

Next, let's call this function by specifying the appropriate bands.

```{r}
ndvi = calculate_ndvi(landsat[["NIR"]], landsat[["Red"]])
```

We obtained a result that falls within the range of -1 to 1. The higher values
often correspond to denser or more vigorous vegetation, while values near or
below 0 may indicate bare soil, artificial surfaces, or water bodies.

Let's now create visualizations using a custom color palette ranging from
red (low values) to green (high values). We can generate this palette by
passing the main colors to the `colorRampPalette()` function and the
number of interpolated intervals between them.

```{r}
# divergent color palette
ndvi_colors = colorRampPalette(c("#8B0000", "#F5DEB3", "#006400"))
plot(ndvi, main = "NDVI", col = ndvi_colors(20), range = c(-1, 1))
```

Forests and croplands are marked in green, while urban areas and bodies of
water are marked in yellow.

## Thresholding

We can use a threshold on the NDVI raster to create a simple binary mask
showing areas covered by vegetation. We will use 0.4 as the threshold value.

```{r}
veg = ndvi > 0.4
plot(veg, col = c("white", "darkgreen"), main = "Vegetation")
```

## Simple Classification

A more complex approach is based on classification rules. To reclassify raster
values based on conditions, we use the `classify()` function. It requires
a matrix consisting of three columns. The first and second columns contain
the starting and ending range of raster values, and the third column contains
the new category value (ID).

```{r}
# reclassification matrix
rcl = matrix(
  c(-1.0, 0.2, 1,   # Non-vegetated / Water
     0.2, 0.4, 2,   # Sparse vegetation
     0.4, 0.6, 3,   # Moderate vegetation
     0.6, 1.0, 4),  # Dense vegetation
  ncol = 3, byrow = TRUE
)
```

```{r}
# raster value reclassification
ndvi_class = classify(ndvi, rcl)
```

The result is a new raster consisting of IDs from 1 to 4. Let's replace these
IDs with category names. We need to create a data frame with two columns.
The first column contains IDs, and the second contains labels. Then, using
the `levels()` function, we can assign these labels to their corresponding IDs.

```{r}
df = data.frame(
  id = rcl[, 3],
  labels = c("Non-vegetated", "Sparse", "Moderate", "Dense")
)

# assign labels to IDs 
levels(ndvi_class) = df
```

Let's take a look at how these categories look on the map.

```{r}
col_class = c("#e2e8f0", "#fefae0", "#78c679", "#065f46")
plot(ndvi_class, col = col_class, main = "Vegetation Density")
```

We can also count the number of occurrences of each category using the `freq()`
function.

```{r}
# frequency table
freq(ndvi_class, bylayer = FALSE)
```

## Focal Operations

The created map shows a "salt-and-pepper" effect, in which incorrect categories
appear within homogeneous areas. This occurs because classification is applied
independently to each cell without considering spatial context, making it
sensitive to local spectral variability. This can be corrected by using focal
operations, which analyze the local neighborhood of cells. In our case, we
can use a modal filter in the `focal()` function to select the most frequently
appearing categories within a moving window of a specified size. Let's assume
the window size is 5 x 5 cells (`w = 5`).

```{r}
# map smoothing using modal filter
ndvi_class_smooth = focal(ndvi_class, w = 5, fun = "modal", expand = TRUE)
```

Let's look at the smoothed map.

```{r}
plot(ndvi_class_smooth, col = col_class, main = "Vegetation Density")
```

## Exporting Results

We can save the results of our analysis as a raster using the `writeRaster()`
function and specifying the save path along with the file extension.
GeoTIFF files are saved with LZW compression by default.

```{r}
#| eval: false

# save raster to disk
writeRaster(ndvi, "NDVI.tif")
```

For a raster with categories, we can define the data type as an 8-bit unsigned
integer (`datatype = "INT1U"`) instead of a floating-point number. Since our
classes only range from 1 to 4, this significantly reduces the file size
on disk.

```{r}
#| eval: false

writeRaster(ndvi_class, "NDVI_class.tif", datatype = "INT1U")
```

# Exercise

1. Create a false color composition consisting of the following bands in this
order: "NIR", "Red", "Green".
2. Calculate the normalized difference water index (NDWI) for water detection.
Next, use thresholding to delineate the water areas and save the result to
disk in TIF format. To calculate NDWI, use the following formula:

$$
\text{NDWI} = \frac{\text{Green} - \text{NIR}}{\text{Green} + \text{NIR}}
$$

