---
title: "Vector data analysis in R"
author: "Krzysztof Dyba"
format:
  html:
    toc: true
    toc-title: " "
    embed-resources: true
---

```{r}
#| message: false
#| warning: false

# load the package
library("terra")

# prevent scientific notation
options(scipen = 8)
```

# Introduction

In the `data` folder, there are two vector files with the `.gpkg` extension
(*geopackage*). They represent the districts of Warsaw and parks as polygon
objects. The data was downloaded from [OpenStreetMap](https://www.openstreetmap.org/)
using the [osmdata](https://github.com/ropensci/osmdata) package. The purpose
of this analysis is to identify the accessibility of parks in the urban
environment.

Let's start by loading these two files (`districts.gpkg` and `parks.gpkg`)
using the `vect()` function. We must specify the paths to these files in
the function.

```{r}
# load vector layer
districts = vect("data/districts.gpkg")
districts
```

We can access the `name` attribute using the dollar sign (`districts$name`),
which returns a vector, or using double square brackets (`districts[["name"]]`),
which returns a data frame.

```{r}
# names of districts in Warsaw
districts$name
```

Now let's load the parks.

```{r}
# load vector layer
parks = vect("data/parks.gpkg")
parks
```

After loading vector data, it is good practice to check that the geometry is
valid. The `is.valid()` function is designed for this purpose. If a layer
consists of multiple objects, it is advisable to use the additional `all()`
function, which returns whether the geometries of all objects are valid.
If the geometry is incorrect, you can try to fix it using the `makeValid()`
function, although this does not always work.

```{r}
# check geometry validity
all(is.valid(districts))
all(is.valid(parks))
```

For simple visualizations, we can use the `plot()` function. If we want to
overlay several layers, we must use the `add = TRUE` argument. The first layer
is displayed as the base layer, and subsequent layers are added on top of it.

```{r}
plot(districts, main = "Parks in Warsaw")
plot(parks, add = TRUE, col = "darkgreen")
```

The `parks` object contains some park names. Let's check how many parks have
no name (`NA`). We can check for missing values using the `is.na()` function
on the `name` attribute. Then, we can count the occurrences using the `table()`
function.

```{r}
# is park name missing?
table(is.na(parks$name))
```

# Vector Operations

## Polygon Statistics

Let's calculate the area and perimeter of polygons. The `expanse()` function
is for calculating area, while the `perim()` function is for calculating
perimeter. For area, we can specify the unit of the result (the `unit` argument).

```{r}
# calculate the area
parks$area_m2 = expanse(parks, unit = "m")
summary(parks$area_m2)
```

```{r}
# calculate the perimeter
parks$perim = perim(parks)
summary(parks$perim)
```

Now that we have the results, we can sort the objects from largest to smallest
area. Objects of the `SpatVector` class behave similarly to data frames, so
we can use the `order()` function to sort them.

```{r}
# sort by area
largest_parks = parks[order(parks$area_m2, decreasing = TRUE), ]
```

We can display the results as a classic data frame after converting them using
`as.data.frame()` function.

```{r}
# convect SpatVector to data frame
largest_parks = as.data.frame(largest_parks)
head(largest_parks)
```

In the further part of the analysis, let’s exclude small parks with an area of
less than 2,000 $m^2$. We need to use square brackets with the appropriate
condition on the `area_m2` attribute.

```{r}
# data filtering
parks_clean = parks[parks$area_m2 > 2000]
```

## Spatial Relationships

Based on the map, we can conclude that the spatial distribution of parks is not
uniform and that some parks are located outside the administrative boundaries
of Warsaw. Spatial relationships define how objects are related to each other
in terms of location, distance, and interaction in space.

So let's check which parks are located in specific districts. To do this, we
need to use the `relate()` function, which will check the selected relationship
between all objects. The result will be a matrix of logical values. The relation
we are interested in is `intersects`.

```{r}
# check if objects intersect
intersect_mat = relate(districts, parks_clean, relation = "intersects")
# View(intersect_mat)
```

Next, for each district, we can count the total number of parks. The districts
are defined in rows, so we need to use the `rowSums()` function.

```{r}
stats = data.frame(name = districts$name, count = rowSums(intersect_mat)) 
head(stats)
```

## Availability Analysis

```{r}
buffers_300 = buffer(parks_clean, width = 300)
buffers_1000 = buffer(parks_clean, width = 1000)
```

```{r}
buffers_300 = aggregate(buffers_300)
buffers_1000 = aggregate(buffers_1000)
```

```{r}
buffers_300 = crop(buffers_300, districts)
buffers_1000 = crop(buffers_1000, districts)
```

```{r}
city_area = sum(expanse(districts))

stats = data.frame(
  threshold = c("300 m", "1000 m"),
  covered_area_m2 = c(expanse(buffers_300), expanse(buffers_1000))
)
stats$pct_covered = stats$covered_area_m2 / city_area * 100

stats
```

```{r}
no_access = erase(aggregate(districts), buffers_1000)
```

```{r}
expanse(no_access) / city_area * 100
```

## Proximity Analysis

```{r}
pt = vect(cbind(21.007, 52.222), crs = "EPSG:4326")
```

```{r}
idx = nearby(pt, parks_clean, k = 1, centroids = TRUE)[, 2]
```

```{r}
distance(pt, parks_clean[idx, ], unit = "m")
```

```{r}
cent = centroids(parks_clean[idx, ])
```

```{r}
crds(cent)
```

```{r}
plot(aggregate(districts), legend = FALSE)
plot(pt, add = TRUE, col = "red")
plot(cent, add = TRUE, col = "darkgreen")
add_legend("topright", legend = c("Park", "Faculty"), pch = 19,
           col = c("darkgreen", "red"))
```

## Exporting Results

```{r}
# project()
# writeVector()
```

# Exercise

1. For each district, calculate its centroid. Then, for each centroid, find
the nearest park and calculate the distances.


