---
title: "Vector data analysis in R"
author: "Krzysztof Dyba"
format:
  html:
    toc: true
    toc-title: " "
    embed-resources: true
---

```{r}
#| message: false
#| warning: false

# load the package
library("terra")

# prevent scientific notation
options(scipen = 8)
```

# Introduction

In the `data` folder, there are two vector files with the `.gpkg` extension
(*geopackage*). They represent the districts of Warsaw and parks as polygon
objects. The data were downloaded from [OpenStreetMap](https://www.openstreetmap.org/)
using the [osmdata](https://github.com/ropensci/osmdata) package. The purpose
of this analysis is to identify the accessibility of parks in the urban
environment.

Let's start by loading these two files (`districts.gpkg` and `parks.gpkg`)
using the `vect()` function. We must specify the paths to these files in
the function.

```{r}
# load vector layer
districts = vect("data/districts.gpkg")
districts
```

We can access the `name` attribute using the dollar sign (`districts$name`),
which returns a vector, or using double square brackets (`districts[["name"]]`),
which returns a data frame.

```{r}
# names of districts in Warsaw
districts$name
```

Now let's load the parks.

```{r}
# load vector layer
parks = vect("data/parks.gpkg")
parks
```

After loading vector data, it is good practice to check that the geometry is
valid. The `is.valid()` function is designed for this purpose. If a layer
consists of multiple objects, it is advisable to use the additional `all()`
function, which returns whether the geometries of all objects are valid.
If the geometry is invalid, you can try to fix it using the `makeValid()`
function, although this does not always work.

```{r}
# check geometry validity
all(is.valid(districts))
all(is.valid(parks))
```

For simple visualizations, we can use the `plot()` function. If we want to
overlay several layers, we must use the `add = TRUE` argument. The first layer
is displayed as the base layer, and subsequent layers are added on top of it.

```{r}
plot(districts, main = "Parks in Warsaw")
plot(parks, add = TRUE, col = "darkgreen")
```

The `parks` object contains some park names. Let's check how many parks have
no name (`NA`). We can check for missing values using the `is.na()` function
on the `name` attribute. Then, we can count the occurrences using the `table()`
function.

```{r}
# is park name missing?
table(is.na(parks$name))
```

# Vector Operations

## Polygon Statistics

Let's calculate the area and perimeter of polygons. The `expanse()` function
is for calculating area, while the `perim()` function is for calculating
perimeter. For area, we can specify the unit of the result (the `unit` argument).

```{r}
# calculate the area
parks$area_m2 = expanse(parks, unit = "m")
summary(parks$area_m2)
```

```{r}
# calculate the perimeter
parks$perim_m = perim(parks)
summary(parks$perim_m)
```

Now that we have the results, we can sort the objects from largest to smallest
area. Objects of the `SpatVector` class behave similarly to data frames, so
we can use the `order()` function to sort them.

```{r}
# sort by area
largest_parks = parks[order(parks$area_m2, decreasing = TRUE), ]
```

We can display the results as a classic data frame after converting them using
`as.data.frame()` function.

```{r}
# convert SpatVector to data frame
largest_parks = as.data.frame(largest_parks)
head(largest_parks)
```

In the following part of the analysis, let’s exclude small parks with an area
of less than 2,000 $m^2$. We need to use square brackets with the appropriate
condition on the `area_m2` attribute.

```{r}
# data filtering
parks_clean = parks[parks$area_m2 > 2000, ]
```

## Spatial Relationships

Based on the map, we can conclude that the spatial distribution of parks is not
uniform and that some parks are located outside the administrative boundaries
of Warsaw. Spatial relationships define how objects are related to each other
in terms of location, distance, and interaction in space.

So let's check which parks are located in specific districts. To do this, we
need to use the `relate()` function, which will check the selected relationship
between all objects. The result will be a matrix of logical values. The
relationship we are interested in is `intersects`.

```{r}
# check if objects intersect
intersect_mat = relate(districts, parks_clean, relation = "intersects")
# View(intersect_mat)
```

Next, for each district, we can count the total number of parks. The districts
are defined in rows, so we need to use the `rowSums()` function.

```{r}
# count parks by districts
district_stats = data.frame(name = districts$name, count = rowSums(intersect_mat)) 
head(district_stats)
```

Note that parks intersecting each district are counted. If a park is located
in two districts, it will be included in both.

## Availability Analysis

Let’s now assess the accessibility of parks. We can consider two scenarios:

1. Parks with good accessibility within 300 m (average walking time ~4 minutes).
2. Parks with moderate accessibility within 1000 m (average walking time ~15
minutes).

To perform this analysis, we need to use buffers. A buffer is a zone of a
specified width created around a selected geographic feature (point, line,
or polygon). This allows us to determine what lies within a specified range
of the location of interest. However, this distance is measured in a straight
line (it does not show the real walking distance along the street network).
To create the buffer, we will use the `buffer()` function and specify the
range in meters (the `width` argument).

```{r}
# generate buffers
buffers_300 = buffer(parks_clean, width = 300)
buffers_1000 = buffer(parks_clean, width = 1000)
```

We get 383 buffers (the same number as the number of input objects). These
buffers largely overlap. Therefore, we can dissolve them into a single
(multi)polygon using the `aggregate()` function.

```{r}
# combine overlapping geometries
buffers_300 = aggregate(buffers_300)
buffers_1000 = aggregate(buffers_1000)
```

Some of the parks and their buffers extend beyond the city boundaries. We can
clip them exactly to the city boundaries using the `crop()` function.

```{r}
# clip buffers to the city boundary
buffers_300 = crop(buffers_300, districts)
buffers_1000 = crop(buffers_1000, districts)
```

```{r}
plot(districts, main = "Park Accessibility", border = NA, legend = FALSE)
plot(buffers_1000, col = "lightblue", border = NA, add = TRUE)
plot(buffers_300, col = "darkblue", border = NA, add = TRUE)
plot(districts, lwd = 2, add = TRUE)
add_legend("topright", legend = c("4 min", "15 min"), pch = 15,
           col = c("darkblue", "lightblue"))
```

We can calculate the total area covered by these buffers and the ratio to the
area of the entire city as a percentage.

```{r}
# the total area of the city
city_area = sum(expanse(districts))

# buffer area
coverage_stats = data.frame(
  threshold = c("300 m", "1000 m"),
  covered_area_m2 = c(expanse(buffers_300), expanse(buffers_1000))
)
coverage_stats$pct_covered = coverage_stats$covered_area_m2 / city_area * 100

coverage_stats
```

We can also identify the areas where access to parks is more difficult
(distances greater than 1,000 m). We can do this by subtracting two vector
layers using the `erase()` function.

```{r}
# vector layers difference
bad_access = erase(aggregate(districts), buffers_1000)
```

```{r}
plot(districts, main = "Park Accessibility", col = "lightblue", border = NA,
     legend = FALSE)
plot(bad_access, col = "red", border = NA, add = TRUE)
plot(districts, lwd = 2, add = TRUE)
add_legend("topright", legend = c("<15 min", ">15 min"), pch = 15,
           col = c("lightblue", "red"))
```

## Proximity Analysis

Finally, let's find the nearest park located next to our faculty. The
coordinates of our faculty are $X = 21.007$ (longitude) and $Y = 52.222$
(latitude) in the geographic coordinate system (`EPSG:4326`).

Let's create a point that represents these geographic coordinates. The simplest
way is to create a matrix with two columns. The first column will represent
longitude, and the second will represent latitude. Next, pass the matrix to
the `vect()` function and define the coordinate system in the `crs` argument.

```{r}
# create point (SpatVector) from coordinates
pt = cbind(21.007, 52.222)
pt = vect(pt, crs = "EPSG:4326")
```

To find the nearest park, we will use the `nearby()` function, which searches
for `k` nearest neighbors (default 1). The result is a matrix containing the
indices of the nearest objects.

```{r}
# find the nearest park
idx = nearby(pt, parks_clean, centroids = FALSE)
idx = idx[, 2]
```

Next, using the `distance()` function, we can calculate the distance between
the faculty and this park. By default, the distance is returned in meters.

```{r}
# calculate distance
distance(pt, parks_clean[idx, ], unit = "m")
```

The distance between the faculty and the nearest park is ~411 meters.

In the `nearby()` function, you may have noticed that we used the `centroids`
argument set to `FALSE`. By default, it is set to `TRUE`, which means that the
centroid will be used to calculate distances for polygons. A centroid is the
geometric center of a spatial object, represented as a single point. Therefore,
the difference in distance to the boundary of the polygon and to its centroid
will be different. In our examples, we calculated the distance to the polygon
boundary.

The `centroids()` function is used to calculate centroids. Let’s find out where
the centroid of the selected park is located.

```{r}
# calculate centroids
cent = centroids(parks_clean[idx, ])
```

```{r}
plot(parks_clean[idx, ], main = "Park", legend = FALSE)
plot(cent, add = TRUE, col = "blue", cex = 1.4)
add_legend("topleft", legend = "Centroid", pch = 19, col = "blue")
```

We can determine the coordinates of spatial objects (including points) using
the `crds()` function.

```{r}
# centroid coordinates
crds(cent)
```

## Reproject 

We can transform (reproject) the coordinate system using the `project()` function.
Simply specify the target coordinate system, preferably by providing the EPSG
code. A list of codes for specific areas based on the map can be found on the
[CRS Explorer](https://crs-explorer.proj.org/) website. In our example, we will
use the *Polish National Geodetic Coordinate System 1992* as the target
coordinate system, whose code is `EPSG:2180`.

```{r}
# reproject CRS
buffers_300_epsg2180 = project(buffers_300, "EPSG:2180")
```

After the above transformation, the coordinates are now expressed in meters
(they were previously in degrees). 

## Exporting Results

Let's export the results to a vector file to disk. The most common format is
geopackage (`.gpkg`). The `writeVector()` function is needed to save the data.

```{r}
#| eval: false

# save buffers to disk
writeVector(buffers_300, "buffers_300.gpkg")
```

# Exercise

1. For each district, calculate its centroid. Then, for each centroid, find
the nearest park and calculate the distances.

