---
title: "Spatial data visualization in R"
author: "Krzysztof Dyba"
format:
  html:
    toc: true
    toc-title: " "
    embed-resources: true
---

```{r}
#| message: false
#| warning: false

# load the packages
library("terra")
library("ggplot2")
library("tidyterra")
library("ggspatial")
```

# Introduction

The `SpatRaster` and `SpatVector` classes have `plot()` methods for creating
quick and simple visualizations. However, it is possible to create more
advanced visualizations using the [tidyterra](https://dieghernan.github.io/tidyterra/)
package, which provides seamless integration with the [ggplot2](https://ggplot2.tidyverse.org/)
package. The purpose of this part of the workshop is to prepare visualizations
of raster and vector data.

# Visualization 

The two essential functions are:

- `geom_spatraster()` for visualizing raster data.
- `geom_spatvector()` for visualizing vector data.

The remaining plot elements (e.g., scales, labels, themes) are identical to
those in `ggplot2`. Therefore, from a technical perspective, the process of
visualizing non-spatial and spatial data is similar in `ggplot2`.

## Raster Data

### Load and Preprocess Data

Let's start by loading and preparing the raster data, as we did before. In
addition, let's reduce the spatial resolution (downsample) to speed up the
rendering process (nevertheless, `tidyterra` does this automatically as well).

```{r}
files = list.files("data/landsat", pattern = "\\.TIF$", full.names = TRUE)
landsat = rast(files)
names(landsat) = c("Blue", "Green", "Red", "NIR")
```

```{r}
scale_factor = 0.0000275
offset = -0.2
landsat = landsat * scale_factor + offset
```

```{r}
# downsample raster to 60 m
landsat = aggregate(landsat, fact = 2, fun = "mean")
```

```{r}
landsat = clamp(landsat, lower = 0, upper = 1, values = TRUE)
```

### Single Band

Let's create a visualization of a single band (NIR). The code syntax is identical
to that used for non-spatial data. The only difference is that we need to define
the raster layer using `geom_spatraster()` function. We can change the color
scale using `scale_fill_gradient()` function and defining the colors.

```{r}
#| message: false

ggplot() +
  geom_spatraster(data = landsat[[4]]) +
  scale_fill_gradient(low = "black", high = "white") +
  labs(title = "NIR Band", fill = "Reflectance") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

### Scale and North Arrow

The north arrow and the scale are integral parts of spatial data visualization.
To add them, we need to use the `ggspatial` package, which includes the
`annotation_north_arrow()` and `annotation_scale()` functions.

```{r}
#| message: false

p = ggplot() +
  geom_spatraster(data = landsat[[4]]) +
  scale_fill_gradient(low = "black", high = "white") +
  annotation_north_arrow(
    location = "tr", height = unit(1, "cm"), width = unit(1, "cm"),
    pad_x = unit(1, "cm"), pad_y = unit(1, "cm")) +
  annotation_scale(
    location = "bl", pad_x = unit(1, "cm"), pad_y = unit(1, "cm"),
    line_col = "white", text_col = "white", text_cex = 1) +
  labs(title = "NIR Band", fill = "Reflectance") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
p
```

### Faceting

If we have several raster layers, we can arrange them side by side in a single
panel using faceting (`facet_wrap()` function).

```{r}
#| message: false

ggplot() +
  geom_spatraster(data = landsat[[1:3]]) +
  scale_fill_gradient(low = "black", high = "white") +
  facet_wrap(~lyr) +
  labs(fill = "Reflectance") +
  theme_void() +
  theme(
    legend.position = "bottom"
  )
```

### RGB Composition

We can also prepare a composition of selected bands using the `geom_spatraster_rgb()`
function.

```{r}
#| message: false

ggplot() +
  geom_spatraster_rgb(data = landsat, r = 3, g = 2, b = 1, stretch = "lin") +
  labs(title = "RGB Composition") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

### Export Visualization

Obviously, we can save the created visualization to disk using the `ggsave()`
function. We can also specify the exact dimensions (`width`, `height`, `units`)
and resolution (`dpi`) of the image.

```{r}
#| eval: false

ggsave("NIR.png", p, width = 6, height = 4, units = "in")
```

## Vector Data

### Load Data

To visualize the vector data, we will again use layers representing districts
and parks. In addition, let's calculate the area of each district in square
kilometers.

```{r}
districts = vect("data/districts.gpkg")
parks = vect("data/parks.gpkg")
```

```{r}
districts$area_km2 = expanse(districts, unit = "km")
```

### Choropleth Map

Let's display the calculated areas of the districts on a choropleth map. The
procedure is similar to that for visualizing raster data, except that in this
case we need to use the `geom_spatvector()` function to visualize vector data.
We also need to specify the variable (column) we want to present on the map by
defining it inside `aes()`.

```{r}
ggplot() +
  geom_spatvector(data = districts, aes(fill = area_km2), color = "white") +
  scale_fill_gradient(low = "grey", high = "brown") +
  labs(title = "Area of districts in Warsaw", fill = expression("Area [km"^2*"]")) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

### Combining Layers

Raster and vector layers can be overlaid. As an example, we will combine two
vector layers, but raster layers can be handled in the same way.

```{r}
ggplot() +
  geom_spatvector(data = districts, fill = NA) +
  geom_spatvector(data = parks, fill = "darkgreen", color = NA) +
  labs(title = "Parks in Warsaw") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

### Reprojecting 

As we have already stated, our vector data is in a geographic coordinate system
(`EPSG:4326`) and is expressed in degrees. We can project the map for display
purposes ("on the fly") using the `coord_sf()` function (from `ggplot2`).
The `crs` argument defines the target coordinate reference system, while the
`datum` argument specifies the coordinate system for the grid lines and axis
labels. We can display the coordinates on our map in meters (rather than degrees)
by setting the coordinate system to `EPSG:2180`, which is the local coordinate
system for Poland.

```{r}
ggplot() +
  geom_spatvector(data = districts, fill = NA) +
  geom_spatvector(data = parks, fill = "darkgreen", color = NA) +
  coord_sf(crs = "EPSG:2180", datum = "EPSG:2180") +
  labs(title = "Parks in Warsaw") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

# Exercise

1. Create a visualization of Warsaw’s districts with `tidyterra`. Use the RGB
composition as the base map. Include the scale and a north arrow. Save the
image to disk with the `png` extension.

